---
title: "Class 10 Candy Project"
author: "Pierce Ford (PID:59010464)"
date: "10/29/2021"
output: github_document
---

```{r}
candy_file <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

# Favorite Candy

> Q3. What is your favorite candy in the dataset and what is it’s winpercent 
value?

```{r}
candy["Warheads", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Let's try "skimming" the dataset

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the 
majority of the other columns in the dataset?

The winpercent column seems to be organized on an 100 point scale whereas most 
are on a 0-1 scale (in fact most are binary)

> Q7. What do you think a zero and one represent for the candy$chocolate column?

Zero and one are being used as logicals for "is this candy chocolatey" (0 is not
chocolatey, 1 is chocolatey)

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution isn't perfectly symmetrical.

> Q10. Is the center of the distribution above or below 50%?

The center is slightly below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
choc.win <- candy$winpercent[as.logical(candy$chocolate)]
fruit.win <- candy$winpercent[as.logical(candy$fruity)]

#Is chocolatey candy better than fruity candy?
mean(choc.win) > mean(fruit.win)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(choc.win, fruit.win)
```

> Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(candy$winpercent, decreasing = TRUE),], n=5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

Color!

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
> Q17. What is the worst ranked chocolate candy?

Sixlets.

> Q18. What is the best ranked fruity candy?

Starburst.
```{r}
rownames(candy)
```


```{r}
library(ggrepel)

my_cols[as.logical(candy$fruity)] = "red"

rownames(candy) <- gsub("Õ", "'", rownames(candy))

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

# Corrplot
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

# PCA

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

