---
title: "Machine Learning 1"
author: "Pierce Ford (PID: A59010464)"
date: "10/22/2021"
output: html_document
---

#Clustering Methods

Kmeans clustering in R is done with the `kmeans()` function.
Here we make up some data to test and learn with.

```{r}
tmp <- c(rnorm(30, 3), rnorm(30, -3))
#Make a two column dataset that includes tmp and tmp reversed, to make two clear
#groups of points
data <- cbind(tmp, rev(tmp))
plot(data)
```

Run `kmeans()` set k (centers) to 2 and nstart to 20. The thing with Kmeans is 
you have to tell it how many clusters you want.

```{r}
km <- kmeans(data, centers=2, nstart=20)
km
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What 'component of your result object details cluster 
assignment/membership?

```{r}
km$cluster
```

> Q. What 'component of your result object details cluster cluster center?

```{r}
km$centers
```


> Q. Plot x colored by the kmeans cluster assignment and add cluster centers as 
blue points?

```{r}
plot(data, col=km$cluster)
points(km$centers, col="blue", pch=15)
```

#Hierarchal Clustering

We will use the `hclust()`function on the same data as before and see how this 
method works.

```{r}
hc <- hclust(dist(data))
hc
```

Hclust has a plot method

```{r}
plot(hc, cex=0.7)
```

To find our membership vector we need to "cut" the tree and for this we use the 
`cutree()` function and tell it the height to cut at.

```{r}
cutree(hc, h=7)
```

We can also use `cutree()` and state the number of k cluster we want.

```{r}
grps <- cutree(hc, k=2)
#plot with hc clusters as color
plot(data, col=grps)
```

#Principal Component Analysis (PCA)