---
title: "Class 17 Vaccination Mini Project"
author: 'Pierce Ford (PID: A59010464)'
date: "11/24/2021"
output:
  pdf_document:
    latex_engine: xelatex
---

# Background

In this before Thanksgiving class when many of our class mates are traveling 
let's have a look at COVID-19 vaccination rates around the state.

We get vaccination rate data from CA.gov here:
https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

## Import Data

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. How many entries do we have?

```{r}
nrow(vax)
```

We can use the **skimr** package and the `skim()` function to get a quick 
overview of the structure of this dataset.

```{r}
skimr::skim(vax)
```

Notice that one of these columns is a date column. Working with time and dates 
gets annoying quickly. We can use the **lubridate** package to make this easy.

```{r}
library(lubridate)

#example lubridate function
today()
```

> Q2. How many days since the first entry in the dataset?

```{r}
vax$as_of_date[1]

#Change this to a date object and do math with today
today() - ymd(vax$as_of_date[1])

#This works, so let's update the whole column
vax$as_of_date <- ymd(vax$as_of_date)

#Test that overwrite worked
today() - vax$as_of_date[1]
```

> Q3. When was the dataset last updated? What is the last date in this dataset 
and how many days have passed since then?

> Q4. How many days does the dataset span?

```{r}
#Last day
last_day <- vax$as_of_date[nrow(vax)]
last_day

#Days since last day
today() - last_day

#Span of data
last_day - vax$as_of_date[1]
```

> Q5. How many different ZIP code areas are in this dataset? 

```{r}
#Saving the zip codes as factors might be useful in describing the data "by ZIP code" later
zip_factors <- as.factor(vax$zip_code_tabulation_area)
#Print how many different "levels" (ZIP codes) there are
length(levels(zip_factors))
```

To work with ZIP codes, we can use **zipcodeR**

```{r}
library(zipcodeR)

#example of using zipcodeR to pull up real world info about the ZIP codes
reverse_zipcode(c('92037', "92109"))
```

## Focus in on San Diego County

Changing track from ZIP codes for a moment. We want to subset the full vax data 
down to just San Diego County.

```{r}
#Subset and check with base R
vax_san_diego_base <- vax[vax$county == "San Diego",]
nrow(vax_san_diego_base)
head(vax_san_diego_base)
```

Subsetting can get tedious and complicated when we have multiple criteria, so we 
can use the **dplyr** package.

```{r}
library(dplyr)
```

We will use the `filter()` function to do our subsetting since it is more 
robust. Focus in on San Diego County as before.

```{r}
#Get San Diego entries and check
sd <- filter(vax, county=="San Diego")
nrow(sd)
```

More complicated subsetting:

```{r}
#Filter by county AND population and check
sd.20 <- filter(vax, county=="San Diego",
                age5_plus_population > 20000)
nrow(sd.20)
```

> Q6. What is the average vaccination rate of San Diego County as of yesterday?

```{r}
#Filter and check
sd.now <- filter(vax, county=="San Diego", 
                 as_of_date=="2021-11-23")
head(sd.now)

#Summary of vaccination percent, including mean
summary(sd.now$percent_of_population_fully_vaccinated)
```

> Q7. Make a histogram of these values.

```{r}
#Base R
hist(sd.now$percent_of_population_fully_vaccinated)
```

However, this method skews for small counties (strong per-person effect on 
percentages).

> Q8. What is the population of the 92037 ZIP code (UCSD) area?

> Q9. What is the average vaccination value for this UCSD/La Jolla ZIP code 
area?

```{r}
#Filter and view
lj <- filter(sd.now, zip_code_tabulation_area=="92037")
lj

#Population and vaccination rate
lj$age5_plus_population
lj$percent_of_population_fully_vaccinated
```

> Q10. What about my Thanksgiving ZIP code?

```{r}
home.vax <- filter(vax, zip_code_tabulation_area=="92882", 
                 as_of_date=="2021-11-23")
#Percent Vaxxed
home.vax$age5_plus_population
home.vax$percent_of_population_fully_vaccinated
```

## Time Course for Here and Home

Start with time course for 92037.

```{r}
#load ggplot
library(ggplot2)

#filter for here (92037)
here.by_date <- filter(vax, zip_code_tabulation_area=="92037")

#plot here
ggplot(here.by_date) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) + 
  geom_point() +
  geom_line(group=1) +
  labs(x="Date", y="Percent Vaccinated")
```

Repeat for Home (92882).

```{r}
#filter for home (92882)
home.by_date <- filter(vax, zip_code_tabulation_area=="92882")

#plot here
ggplot(home.by_date) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) + 
  geom_point() +
  geom_line(group=1) +
  labs(x="Date", y="Percent Vaccinated")
```

Let's compare across all of San Diego county.

> Q11. How many ZIP codes pass this filter?

```{r}
#filter for large zip codes
sd.large <- filter(vax, county=="San Diego",
                       age5_plus_population >= 36144)

#how many ZIPs
length(levels(as.factor(sd.large$zip_code_tabulation_area)))

#plot sd.large
ggplot(sd.large) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated, col=as.factor(zip_code_tabulation_area), group=zip_code_tabulation_area) + 
  geom_line(alpha=0.5) +
  labs(x="Date", y="Percent Vaccinated")
```

Repeat for all of California.

```{r}
#filter for large zip codes
ca.large <- filter(vax, age5_plus_population >= 36144)

#how many ZIPs
length(levels(as.factor(ca.large$zip_code_tabulation_area)))

#plot ca.large
ggplot(ca.large) + 
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated, col=as.factor(zip_code_tabulation_area), group=zip_code_tabulation_area) + 
  geom_line(alpha=0.2) +
  labs(x="Date", y="Percent Vaccinated") +
  theme(legend.position = "none") +
  geom_hline(yintercept = 0.67)
```

What is the mean across the state for these large pop ZIP codes?

```{r}
#Filter for yesterday
ca.large.now <- filter(ca.large, as_of_date=="2021-11-23")

#Summary of vaccination percent, including mean
summary(ca.large.now$percent_of_population_fully_vaccinated)
```

